---
title: "Covid-19 study"
author: "Kevin Juandi"
date: "2024-03-22"
output: html_document
---

# Installing Libraries

```{r, eval=FALSE, echo=TRUE}
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("repr")
```

# Loading Libraries

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(repr, warn.conflicts = FALSE)
options(repr.plot.width=10, repr.plot.height=8)
options(dplyr.summarise.inform = FALSE)
options(warn = -1)
```


# Loading the Data

Data Location: https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series

This data is the list of COVID-19 cases by country and province. I will use three files consisting of data for confirmed cases, deaths, and recoveries.

```{r, echo=TRUE}
death_global_data =
  read.csv(file = paste(
    'https://github.com/CSSEGISandData/COVID-19',
    'tree/master/csse_covid_19_data/csse_covid_19_time_series',
    'time_series_covid19_deaths_global.csv',
    sep = '/'))
confirmed_global_data =
  read.csv(file = paste(
    'https://github.com/CSSEGISandData/COVID-19',
    'tree/master/csse_covid_19_data/csse_covid_19_time_series',
    'time_series_covid19_confirmed_global.csv',
    sep = '/'
  ))
recovered_global_data =
  read.csv(file = paste(
    'https://github.com/CSSEGISandData/COVID-19',
    'tree/master/csse_covid_19_data/csse_covid_19_time_series',
    'time_series_covid19_recovered_global.csv',
    sep = '/'
  ))

```

# Cleaning the data
```{r, echo=TRUE}
tidy_data <- function(data) {
  result <- data %>%
    pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
                 names_to = 'Date',
                 values_to = 'Cases') %>%
    select(-c(Lat, Long)) %>%
    mutate(Date = mdy(substring(Date, 2, length(Date)))) %>%
    group_by(Date) %>%
    summarise(
      Cases = sum(Cases),
    )
}

death_global_data_tidy <- tidy_data(death_global_data) %>%
  mutate(Deaths = Cases) %>%
  select(-Cases)
confirmed_global_data_tidy <- tidy_data(confirmed_global_data) %>%
  mutate(Confirmed = Cases) %>%
  select(-Cases)
recovered_global_data_tidy <- tidy_data(recovered_global_data) %>%
  mutate(Recovered = Cases) %>%
  select(-Cases)

global <- death_global_data_tidy %>%
  full_join(confirmed_global_data_tidy, by = 'Date') %>%
  full_join(recovered_global_data_tidy, by = 'Date')

```

Here I've converted the date columns to rows and then grouped by date. I've also added a column for the number of cases.
Three datasets were joined together to create a single dataset.