---
title: "NYPD shooting report data visualization and analysis"
author: "Kevin Juandi"
date: "2024-03-16"
output: html_document
---

# Installing Libraries

```{r, eval=FALSE, echo=TRUE}
#install.packages("tidyverse")
#install.packages("lubridate")
#install.packages("ggplot2")
#install.packages("repr")
```

# Loading Libraries

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse, warn.conflicts = FALSE)
library(lubridate, warn.conflicts = FALSE)
library(ggplot2, warn.conflicts = FALSE)
library(repr, warn.conflicts = FALSE)
options(repr.plot.width=10, repr.plot.height=8)
options(dplyr.summarise.inform = FALSE)
```

# Loading Data

```{r, echo=TRUE}
url_NYPD <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
NYPD <- read.csv(url_NYPD)
```

# EDA
Let's start with simple EDA

```{r, echo=TRUE}
head(NYPD, 10)
```

There seems to be a lot of blanks

```{r, echo=TRUE}
sapply(NYPD, function(x) sum(is.na(x)))
```

```{r, echo=TRUE}
summary(NYPD)
```

# Graph Plots

```{r, echo=TRUE}
NYPD_clean <- NYPD %>%
    select(c("OCCUR_DATE","OCCUR_TIME","BORO","PRECINCT",
             "STATISTICAL_MURDER_FLAG","PERP_RACE","VIC_AGE_GROUP","VIC_SEX","VIC_RACE")) %>%
    mutate(OCCUR_DATE = mdy(OCCUR_DATE),
           OCCUR_TIME = hms(OCCUR_TIME),
           STATISTICAL_MURDER_FLAG = as.logical(STATISTICAL_MURDER_FLAG),
           Shootings = 1,
           Year = year(OCCUR_DATE))

NYPD_clean %>%
    ggplot(aes(x = BORO, fill = BORO)) +
    geom_bar() +
    labs(title = "NYPD Shootings Incidents by Borough",
         subtitle = "(2006 - 2021)",
         x = "NYC Boroughs",
         y = "Total Number of Shootings",
         caption = "(Figure - 1)")
```

```{r, echo=TRUE}
NYPD_clean %>%
    ggplot(aes(x = Year)) +
    geom_bar(fill = "lightblue",show.legend = FALSE) +
    labs(title = "NYPD Shootings Incidents by Year",
         x = "Years (2006 - 2021)",
         y = "Total Number of Shootings",
         caption = "(Figure - 2)")
```

```{r, echo=TRUE}
NYPD_year <- NYPD_clean %>%
    group_by(Year) %>%
    summarize(Shootings = sum(Shootings))

NYPD_year %>%
    ggplot(aes(x = as.factor(Year), y = Shootings)) +
    geom_line() +
    geom_point(color = "red") +
    scale_x_discrete(labels = as.character(2006:2021)) +
    labs(
        title = "NYPD Shooting Incidents by Year",
        x = "Years (2006 - 2021)",
        y = "Total Number of Shootings",
        caption = "(Figure - 3)"
    )
```

looks like some entries are missing dates

```{r, echo=TRUE}
NYPD_boro <- NYPD_clean %>%
    group_by(BORO, OCCUR_DATE,Shootings) %>%
    summarize(Shootings = sum(Shootings),
              STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
              .groups = 'drop') %>%
    select(BORO,OCCUR_DATE,Shootings,STATISTICAL_MURDER_FLAG) %>%
    ungroup()

NYPD_boro_year <- NYPD_clean %>%
    mutate(Year = year(OCCUR_DATE)) %>%
    group_by(BORO, Year, Shootings) %>%
    summarize(Shootings = sum(Shootings),
              STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
              .groups = 'drop') %>%
    select(BORO, Year,Shootings,STATISTICAL_MURDER_FLAG) %>%
    ungroup()

NYPD_boro_total <- NYPD_boro_year %>%
    group_by(BORO) %>%
    summarize(Shootings = sum(Shootings))
(7402 + 10365)/ sum(NYPD_boro_total$Shootings)
```

```{r, echo=TRUE}
736/ sum(NYPD_boro_total$Shootings)
```

```{r, echo=TRUE}
NYPD_boro_year %>%
    ggplot(aes(x = Year, y = Shootings,color = BORO)) +
    geom_line() +
    geom_point() +
    labs(title = "NYPD Shootings by Borough by Year",
         subtitle = "(2006 - 2021)",
         x = "Year",
         y = "Number of Shootings",
         caption = "(Figure - 4)")
```

It seems like number of population does not reflect the number of incident. Bronx and Manhattan has the same magnitude of population, so does Brooklyn and Queens. I would assume that this is because of the presence of gangs and organized crime syndicates which Bronx and Brooklyn has the reputation for. It also worth to note that Bronx is the poorest borough.

```{r, echo=TRUE}
NYPD_boro %>%
    ggplot(aes(x = OCCUR_DATE, y = Shootings)) +
    geom_line() + scale_x_date(date_labels = "%Y %b") +
    labs(title = "NYPD Shootings Per Day",
         subtitle = "(2006 - 2021)",
         x = "Date",
         y = "Shootings per day",
         caption = "(Figure - 5)")
```

```{r, echo=TRUE}
NYPD_time_year <- NYPD_clean %>%
    mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d")) %>%
    mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
    group_by(Time_year,Shootings) %>%
    summarize(Shootings = sum(Shootings),
              STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
              .groups = 'drop') %>%
    select(Time_year,Shootings,STATISTICAL_MURDER_FLAG) %>%
    ungroup()

NYPD_time_year %>% slice_max(Shootings, n = 2)
```

```{r, echo=TRUE}
NYPD_July_5 <- NYPD_clean %>%
    mutate(Time_year = format(as.Date(OCCUR_DATE), "%m/%d"),
           Hour = hour(OCCUR_TIME)) %>%
    mutate(Time_year = as.Date(Time_year,"%m/%d")) %>%
    filter(Time_year == "2022-07-05") %>%
    group_by(Hour,Shootings) %>%
    summarize(Shootings = sum(Shootings),
              .groups = 'drop')

NYPD_time_year %>%
    ggplot(aes(x = Time_year, y = Shootings)) +
    geom_line() +
    geom_point(data = NYPD_time_year %>% slice_max(Shootings, n = 2),
               aes(color="Most Shootings")) +
    scale_x_date(date_labels = "%b") +
    labs(title = "NYPD Shootings on that Day",
         subtile = "(2006 - 2021)",
         colour = "",
         x = "Day",
         y = "Shootings on that day",
         caption = "(Figure - 6)")
```

```{r, echo=TRUE}
NYPD_time_day <- NYPD_clean %>%
    group_by(OCCUR_TIME,Shootings) %>%
    summarize(Shootings = sum(Shootings),
              STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
              .groups = 'drop') %>%
    select(OCCUR_TIME,Shootings,STATISTICAL_MURDER_FLAG)

NYPD_time_day %>%
    ggplot(aes(x = OCCUR_TIME, y = Shootings)) +
    geom_line() +
    scale_x_time() +
    labs(title = "NYPD Shootings by the Time of Day",
         subtitle = "(2006 - 2021)",
         x = "Time of day (24 Hour Time System)",
         y = "Shootings",
         caption = "(Figure - 7)")
```

```{r, echo=TRUE}
NYPD_time_hour <- NYPD_clean %>%
  mutate(Hour = hour(OCCUR_TIME)) %>%
  group_by(Hour,Shootings) %>%
  summarize(Shootings = sum(Shootings),
            STATISTICAL_MURDER_FLAG = sum(STATISTICAL_MURDER_FLAG),
            .groups = 'drop') %>%
  mutate(Hour2 = Hour^2) %>%
  select(Hour,Shootings,STATISTICAL_MURDER_FLAG, Hour2)
```

```{r, echo=TRUE}
NYPD_time_hour %>%
    ggplot(aes(x = Hour, y = Shootings)) +
    geom_point() +
    stat_smooth(method = "lm", formula = y ~ x + I(x^2), linewidth = 1) +
    labs(title = "NYPD Shootings by Time of Day per Hour",
         subtitle = "(2006-2021)",
         x = " Hours (24 Hour Time System)",
         y = "Shootings that Hour",
         caption = "(Figure - 8)")
```

It's rather unsurprising that we have more incident at night to the wee hours

# Bias

Are there racial bias?

```{r, echo=TRUE}
736/ sum(NYPD_boro_total$Shootings)
```


